---
title: "Taxonomy assignment from BLAST results"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read-in and prep BLAST output file with column names
### For V2 of blast_taxonomy, adjusted to start with merged data file with column names already included. 

```{r}
s <- read.table("~/bryo_arthro/lyon_pilot/working/lyon_pilot_all_blast_merged_labelled.csv", 
                sep = ",", header = TRUE)  ### changed to comma delimited with headers
```

## Give column names to the data.frame

```{r}
colnames(s)[c(1, 3, 4)] <- c("asv_name", "ncbi_accession", 
                               "percent_match")
```

## Fetch higher taxonomy for all matches

We will use the `taxize` package for this task 

```{r}
library(taxize)
```

The basic steps are to 

1. get the unique ID for each accession number using `genbank2uid` like this:

```{r}
#genbank2uid("1509790295")    ### example
```

That gives us the unique ID, in this case `468654`. Then we can

2. Fetch the higher classification with the `classification` function like this:


```{r}
#classification(468654, db = "ncbi")    ### example
```

Note: it is worth running `use_entrez()` to set-up an ENTREZ API key.  Having the API key will allow you to make more calls to the NCBI databases before you hit any limits.

```{r}
#install.packages("rentrez")
library(rentrez)
set_entrez_key("2edc8029dbf711ec8e46c7d66c093d5e4e09")  ### Kylie's API key, stores it for current R session
```


Now we need to write code to extract the accession number from the BLAST results 

```{r}
acc <- s$ncbi_accession   ### set vector

acc <- gsub("gi\\|", "", acc) ### removes "gi|"

acc <- gsub("\\|.*", "", acc) ### removes anything after "|"

s$acc <- acc  ### add these clean accessions back to the BLAST output data.frame

# there are likely duplicate accession numbers (multiple ASVs match to the same
# accession number) so let's make a vector of just the unique ones so we make 
# a smaller request of the NCBI server

acc_unique <- unique(acc)
```


The function `genbank2uid` can work on the entire vector of accession numbers

```{r}
all_uid <- genbank2uid(acc_unique)

# let's combine `all_uid` with `acc_unique` so that we can match accession
# number to UID
acc_uid_hash <- data.frame(uid = unlist(all_uid), 
                           acc = acc_unique)

# there are also likely duplicate UIDs (multiple accessions get the same 
# species) so we remove duplicates here too
uid_unique <- unlist(all_uid) |> unique()
```

Now we need to specify the taxonomic ranks we actually want because `classification` will give us back a bunch of random stuff. We are also going to write a function to clean up the output of `classification`

```{r}
yes_ranks <- c("kingdom", "phylum", "class", "order", 
               "family", "genus", "species")

# helper function to clean output of `classification`
#' @param x is the raw output (a named list) 
#' @param ranks is a character vector providing the desired ranks
#' @returns a data.frame with columns for each of the ranks in `ranks` and a
#'          column for UID

clean_output <- function(x, ranks) {
    tax <- lapply(x, function(d) {
        d <- d[d$rank %in% ranks, ]
        out <- d$name
        
        names(out) <- d$rank
        out <- out[ranks]
        names(out) <- ranks
        
        return(out)
    })
    
    res <- do.call(rbind, tax) |> as.data.frame()
    
    res <- cbind(uid = rownames(res), res)   ### make a column for UID
    
    rownames(res) <- NULL
    
    return(res)
}
```

Now we can run this function on the classification output for all the unique IDs.

```{r}
raw_higher_tax <- classification(uid_unique, db = "ncbi")
higher_tax <- clean_output(raw_higher_tax, yes_ranks)

head(higher_tax)
```

Note with all these functions that query NCBI can potentially error out due to rate limits etc. So if that happens, and keeps happening, we might need to pursue a different solution 

## Reach concensus about classification

First we need to assign taxonomy back to the BLAST output so we can match ASV ID to taxonomic hierarchy.

In lyon_pilot, the data is too large to run, so split data into chunks?

```{r}
#Split s into chunks
split_data <- function(data, chunk_size) {
  split(data, ceiling(seq_len(nrow(data)) / chunk_size))
}
chunk_size <- 1000  # Define chunk size
s_chunks <- split_data(s, chunk_size)

#Merge each chunk 
blast_tax_chunks <- lapply(s_chunks, function(s_chunk) {
  merged_acc <- merge(acc_uid_hash, higher_tax, by = "uid")  
  merge(merged_acc, s_chunk[, setdiff(names(s_chunk), "species")], by = "acc")
})

#Combine chunks
blast_tax <- do.call(rbind, blast_tax_chunks)

#Now retain only the columns we care about
blast_tax <- blast_tax[, c("asv_name", "acc", "percent_match", "uid", 
                           yes_ranks)]


###ANDY'S ORIGINAL if data frames are not too large to exceed memory
#blast_tax <- merge(acc_uid_hash, higher_tax) |> # add taxonomy to acc numbers
#    merge(y = s[, -which(names(s) == "species")]) # add above to BLAST output

# now retain only the columns we care about
#blast_tax <- blast_tax[, c("asv_name", "acc", "percent_match", "uid", 
#                           yes_ranks)]
```

Now we will compute consensus taxonomy. The context of the BLAST results is that the 10 top hits were requested with a minimum match of 85%. We will use the following rules to compute consensus taxonomy:

1. We will set a parameter `sway = 0.5` that is a percentage 
2. Any hit with a percent match greater than `max(percent_match) - sway` will be kept for computing taxonomy, the rest will be discarded
3. The lowest taxonomic level for which there is consensus in the kept hits will be considered the consensus taxonomic assignment 
4. We will set a parameter `min_match = 97` which is the minimum percent match required to make a species-level assignment; so even if all kept hits agree on a species, but their percent match is less than `min_match` we will shift the taxonomic assignment to a higher taxon
5. Each higher taxonomic level will require a minimum percent match that is `level_dec = 2` less than the previous level (e.g. genus requires 95% match, kingdom requires 85% match) 


```{r}
# function to compute consensus taxonomy
#' @param x is a data.frame in the format of `blast_tax` (computed above) but 
#'          which has been subsetted to just one asv
#' @param sway as described above
#' @param min_match as described above
#' @param level_dec as described above

taxon_consensus <- function(x, sway = 0.5, min_match = 97, level_dec = 2) {
  # browser()  
  # calc max percent match
    maxp <- max(x$percent_match)
    
    # retain hits >= maxp - sway
    x <- x[x$percent_match >= maxp - sway, ]
    
    # if only one remains, that's the one
    nhit <- nrow(x)
    if(nhit == 1) {
        the_tax <- x[, c("asv_name", "percent_match", yes_ranks)]
    } else {
        # calculate mean percent match
        mean_match <- mean(x$percent_match)
        
        # boolean, TRUE = yes agree
        agree <- sapply(yes_ranks, function(r) {
            length(unique(x[[r]])) == 1
        })
        
        # keep only ranks for which there is agreement
        if(all(!agree)) { # no ranks agree
          the_tax <- x[1, c("asv_name", "percent_match", yes_ranks)]
          the_tax[yes_ranks] <- NA
          the_tax$percent_match <- mean_match
        } else {
          agree_id <- which(agree) |> max()
          bad_ranks <- yes_ranks[-(1:agree_id)]
          
          # make a single row data.frame of the agreed taxonomy
          the_tax <- x[1, c("asv_name", "percent_match", yes_ranks)]
          the_tax[, bad_ranks] <- NA # make un-agreed ranks = NA
          the_tax$percent_match <- mean_match
        }
    }
    
    # figure out appropriate level given percent match
    cc <- min_match - ((length(yes_ranks):1) - 1) * level_dec # rank cut offs
    supported_id <- which(the_tax$percent_match >= cc) |> max()
    unsupported_ranks <- yes_ranks[-(1:supported_id)]
    
    # trim to appropriate level
    the_tax[, unsupported_ranks] <- NA
    
    return(the_tax)
} 
```

Let's test this function on a few ASVs

```{r}
#asv1 <- blast_tax[blast_tax$asv_name == unique(blast_tax$asv_name)[1], ]
#asv1

#taxon_consensus(asv1)

#asv4 <- blast_tax[blast_tax$asv_name == unique(blast_tax$asv_name)[4], ]
#asv4

#taxon_consensus(asv4)

#asv5 <- blast_tax[blast_tax$asv_name == unique(blast_tax$asv_name)[5], ]
#asv5

#taxon_consensus(asv5)
```

All that looks good to me, so let's run it on everything

```{r}
tax_final <- lapply(unique(blast_tax$asv_name), function(i) {
    this_asv <- blast_tax[blast_tax$asv_name == i, ]
    taxon_consensus(this_asv)
})

tax_final <- do.call(rbind, tax_final)
```


#Taxonomy exploration and visualization
```{r}
arthropoda_count <- sum(tax_final$phylum == "Arthropoda", na.rm = TRUE)
total_count <- nrow(tax_final)  # includes NAs
percentage_arthropoda <- (arthropoda_count / total_count) * 100
percentage_arthropoda

```
##9.12% of ASVs are arthropoda, but how much of the relative abundance is arthropoda (need to calculate using ASV frequency) 

##Bar plot of phylums
```{r}

library("ggplot2")
phylum_counts <- table(tax_final$phylum, useNA = "ifany")   ### Calculate sum
phylum_counts   ### View table

phylum_df <- as.data.frame(phylum_counts)
colnames(phylum_df) <- c("phylum", "count")
phylum_df$percent <- (phylum_df$count / sum(phylum_df$count)) * 100   ### Calculate percent

ggplot(phylum_df, aes(x = reorder(phylum, -percent), y = percent, fill = phylum)) +     ### Plot
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Phylum", y = "Percentage (%)", title = "Percentage of ASV Phyla Types") +
  theme_minimal() +
  theme(legend.position = "none")

```

##Bar plot of tax levels
```{r}
library("ggplot2")
library("tidyverse")
my_colors <- rainbow(50)

kingdom_phylum_tax <- tax_final %>%
  group_by(kingdom, phylum) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(percent = (count / sum(count)) * 100)

ggplot(kingdom_phylum_tax, aes(x = kingdom, y = percent, fill = phylum)) +
  geom_bar(stat = "identity", color = "black") +
  labs(x = "Kingdom", y = "Percentage (%)", 
       title = "Percentage of ASVs per Phyla within each Kingdom") +
  scale_fill_manual(values = my_colors, na.value = "gray80") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

##How much of the relative abundance is arthropoda by ASV frequency?
```{r}
library(tidyr)
library(dplyr)
library(stringr)
library(ggplot2)
my_colors <- rainbow(50)

tax_final_split <- tax_final %>%
    separate(asv_name, into = c("asv_id", "count"), sep = "_count_")

tax_final_split$count <- as.numeric(tax_final_split$count)

tax_final_split$asv_id <- paste0("ASV", str_pad(gsub("ASV", "", tax_final_split$asv_id), width = 4, pad = "0"))

total_reads <- sum(tax_final_split$count)

kingdom_abundance <- tax_final_split %>%
  group_by(kingdom) %>%
  summarize(total_kingdom_count = sum(count)) %>%
    mutate(percentage = (total_kingdom_count / total_reads) * 100)
kingdom_abundance

phylum_abundance <- tax_final_split %>%
  group_by(phylum) %>%
  summarize(total_phylum_count = sum(count)) %>%
    mutate(percentage = (total_phylum_count / total_reads) * 100)
phylum_abundance

#write.csv(phylum_abundance, "phylum_abundance.csv", row.names = FALSE)

kingdom_total_reads <- tax_final_split %>%
  group_by(kingdom) %>%
  summarize(total_kingdom_reads = sum(count))

kingdom_phylum_abundance <- tax_final_split %>%
  group_by(kingdom, phylum) %>%
  summarize(total_phylum_count = sum(count)) %>%
    ungroup()
kingdom_phylum_abundance

ggplot(kingdom_phylum_abundance, aes(x = kingdom, y = total_phylum_count, fill = phylum)) +
  geom_bar(stat = "identity") +  
  labs(x = "Kingdom", y = "Read frequency (%)", title = "Frequency by Kingdom and Phyla") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  
        legend.position = "right")  
```

##Hone in on Metazoa
```{r}
metazoa <- tax_final_split %>%
    filter(kingdom == "Metazoa")
head(metazoa)

total_reads_metazoa <- sum(metazoa$count)

metazoa_phylum_abundance <- metazoa %>%
  group_by(phylum) %>%
  summarize(total_phylum_count = sum(count)) %>%
    mutate(percentage = (total_phylum_count / total_reads_metazoa) * 100)
metazoa_phylum_abundance

ggplot(metazoa_phylum_abundance, aes(x = phylum, y = percentage)) +
  geom_bar(stat = "identity") +  
  labs(x = "Phyla", y = "Read frequency (%)", title = "Frequency of Phyla in Metazoa only") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  
        legend.position = "right")  
```

##Hone in on Arthropoda
```{r}
arthropoda <- tax_final_split %>%
    filter(phylum == "Arthropoda")
head(arthropoda)

# Replace 'your_table_name' with the actual name of your data frame
#arthropoda_asv_class_counts <- arthropoda %>%
#  group_by(class) %>%
#  summarize(asv_count = n_distinct(asv_id)) %>%
#  arrange(desc(asv_count))

# View the result
#print(arthropoda_asv_class_counts)


#write.csv(arthropoda, "arthropoda_lyon_pilot.csv", row.names = FALSE)

total_reads_arthropoda <- sum(arthropoda$count)

arthropoda_class <- arthropoda %>%
  group_by(class) %>%
  summarize(total_class_count = sum(count)) %>%
  mutate(percentage = (total_class_count / sum(total_class_count)) * 100)

#write.csv(arthropoda_class, "arthropoda_class_summary.csv", row.names = FALSE)

ggplot(arthropoda_class, aes(x = class, y = total_class_count)) +
  geom_bar(stat = "identity") +  
  labs(x = "Class", y = "# of reads", title = "Read frequency of Class in Arthropoda only") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  
        legend.position = "right")  

```


##Go to next notebook investigating ASV abundance per sample, habitat type, site, etc.





