---
title: "Comparative Analyses of Arthropoda in Lyon Trial"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in all final data from dada and Lyon Pilot blast taxonomy analysis. Merge.
```{r}
## This is the ASV to taxonomy data
a <- read.table("~/bryo_arthro/lyon_pilot_git/data/arthropoda_lyon_pilot.csv", 
                sep = ",", header = TRUE) 

## This is the sequence to sample data
s <- read.table("~/bryo_arthro/lyon_pilot_git/data/seqtab_nochim.csv", 
                sep = ",", header = TRUE)
s2 <- as.data.frame(t(s))
colnames(s2) <- s2[1,]
s2 <- s2[-1,]

iucn_arthro <- read.table("~/bryo_arthro/iucn_summary.csv", 
                sep = ",", header = TRUE)

## Add asv_id column to s
s2$Sequence <- rownames(s2)
s3 <- s2[, c(ncol(s2), 1:(ncol(s2) - 1))]

##This is the sequence to ASV data
asv <- read.table("~/bryo_arthro/lyon_pilot_git/data/asv_seq_table.csv", 
                sep = ",", header = TRUE)
colnames(asv)[1] <- "asv_id"  ### rename the first column

## We need to integrate all three together
asv_taxa_table <- merge(asv, a, by = "asv_id")
s3_asv_table <- merge(s3, asv[, c("Sequence", "asv_id")], by = "Sequence", all.x = TRUE)
merged <- merge(s3_asv_table, asv_taxa_table, by = "asv_id", all.x = TRUE)
```

## The merge was successful, however, the non-Arthropoda ASVs are still present from the asv and s files. We need to remove those.
```{r}
library(dplyr)

merged_arthro <- merged %>%
  filter(!is.na(Sequence.y))

merged_arthro$Sequence.y <- NULL
colnames(merged_arthro)[colnames(merged_arthro) == "Sequence.x"] <- "sequence"

##Check output
head(merged_arthro)
```

## Convert data to long format and make Frequency column numeric
```{r}
library(reshape2)
library(ggplot2)

merged_arthro_long <- melt(merged_arthro, id.vars = c("asv_id", "sequence", "count", "percent_match", "kingdom", "phylum", "class", "order", "family", "genus", "species"),
                  variable.name = "Sample", value.name = "Frequency")

merged_arthro_long$Frequency <- as.numeric(trimws(merged_arthro_long$Frequency))
#str(merged_arthro_long)   ### check if it worked
```

##Create column of lowest taxonimc ID
```{r}
library(dplyr)

merged_long_id <- merged_arthro_long %>%
  mutate(lowest_taxon = coalesce(species, genus, family, order, class, phylum, kingdom))
```

##Add sample metadata for final file for analyses
```{r}
metadata <- read.table("~/bryo_arthro/lyon_pilot_git/data/lyon_metadata.csv",
                sep = ",", header = TRUE)

merged_long_id_metadata <- merge(merged_long_id, metadata, by = "Sample", all.x = TRUE)

final <- merged_long_id_metadata

write.csv(final, "final.csv", row.names = FALSE)

```


# Now we can begin analyses on the dataset

## Heatmap of ASV frequency for the highest frequency ASVs; too many to display to include lower occurring ASVs
```{r}
library(dplyr)

subset1_final <- final %>%
  mutate(count = as.numeric(count)) %>%   
  filter(count >= 100 & count <= 43976)

ggplot(subset1_final, aes(x = Sample, y = asv_id, fill = Frequency)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(x = "Sample", y = "ASV ID", title = "ASV frequency across samples") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## Or if we want to see a heatmap of the lowest taxons 
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

final_summed_taxon <- final %>%
  group_by(Sample, lowest_taxon) %>%
  summarize(total_count = sum(Frequency, na.rm = TRUE), .groups = "drop")

ggplot(final_summed_taxon, aes(x = lowest_taxon, y = Sample, fill = total_count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(x = "Lowest Taxon", y = "Sample", title = "Heat Map of Lowest Taxons Across Samples") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))

ggplot(final_summed_taxon, aes(x = lowest_taxon, y = Sample, fill = log10(total_count + 1))) + 
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "blue", 
                      name = "Log10(Count)") + 
  labs(x = "Lowest Taxon", y = "Sample", title = "Log Transformed Heat Map of Lowest Taxons Across Samples") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))
```
## Compare final summed taxon to endangered species list
## RESULT = no matches to any species on IUCN list
```{r}
library(dplyr)

iucn_arthro <- iucn_arthro %>%
  select(-assessmentId, -internalTaxonId, -kingdomName, -phylumName, -orderName, -className, -familyName, -genusName, -infraType, -infraAuthority, -infraName)

iucn_arthro <- iucn_arthro %>%
    rename(lowest_taxon = scientificName)

matches <- intersect(names(iucn_arthro), names(final_summed_taxon))
matches ### see if any matches exist, at minimum the cell "lowest_taxon" should show as a match
```

#Make figure for:
## number of taxa by reads
### group site bryophyte + gauze samples and 

```{r}

```




